Topologically Equivariant \\ Artist Model

### 2025-03-27 slides/main.tex
talk
composition overhaul
equivariance + brought up related work + comments
talk version

%\documentclass[xcolor={dvipsnames}, handout]{beamer}
\documentclass[xcolor={dvipsnames}]{beamer}
\setbeamertemplate{footline}[frame number]
\setbeamersize{description width=-\labelsep}

@@

\graphicspath{.figures/}

\usepackage[backend=bibtex, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
\bibliographystyle{plain}
\bibliography{../paper/bibliography.bib}

%\usetheme{ccnycrest}

@@

\begin{document}
\begin{frame}<presentation:1-|handout:0>{About}
    Slides: https://bit.ly/team329
    \begin{block}{Me (twitter/github: story645)}
        \begin{itemize}
            \item nth year grad student (on my 3rd EO)
            \item former adjunct at CCNY, former Digital Fellow
            \item Matplotlib Community Manager \& Core Developer
        \end{itemize}
    \end{block}

    \begin{block}<presentation:2|handout:0>{Project}
        \begin{itemize}
            \item Funded by Chan Zuckerberg Initiative EOSS 1 \& 3
            \item paper rejected by vizweek last spring
            \item work has since gone all in on category theory 
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}<presentation:1|handout:1>
	\titlepage
    Hannah Aizenman, Tom Caswell, Michael Grossberg\\
    %Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
    %External Member: Dr. Marcus Hanwell
\end{frame}

\section{Introduction}
\begin{frame}<presentation:0|handout:0>{What are we doing?}
    \begin{itemize}
        \item develop a model for describing data to graphic transformations
        \item specify a visualization library architecture based on this model
        \item implement functional(ish) components based on this model using ideas from functional programming
    \end{itemize}
\end{frame}

%% maybe pull out the middle for now? move the 3 stages down to construction? 
\begin{frame}<presentation:1-|handout:0>{What do visualization libraries do?}
    \begin{figure}
        \begin{overprint}
            \onslide<1|handout:1>\includegraphics[width=\linewidth]{figures/flow/s2.png}
            \onslide<2|handout:2>\includegraphics[width=\linewidth]{figures/flow/s_vc.png}
            \onslide<3|handout:3>\includegraphics[width=\linewidth]{figures/flow/s_mark.png}
            \onslide<4|handout:4>\includegraphics[width=\linewidth]{figures/flow/s3.png}
        \end{overprint}
    \end{figure}
\end{frame}

\begin{frame}<presentation:1|handout:0>{Matplotlib}
    \begin{figure}
       \includegraphics[width=\linewidth]{figures/flow/artists.png}
    \end{figure}
\end{frame}

\begin{frame}<presentation:1-|handout:0>
    \frametitle<presentation:1-2|handout:0>{Everything is an Artist}
    \frametitle<presentation:3|handout:0>{Transformations Change Artists into Different Coordinates}
    \begin{overprint}
        \onslide<1|handout:0>\includegraphics[scale=.25]{figures/math/sphx_glr_anatomy_001_2_0x.png}
        \onslide<2|handout:0>\includegraphics[width=\textwidth]{figures/math/artist_tree.png}
        \onslide<3|handout:0>\includegraphics[width=\textwidth]{figures/math/transforms.png}
    \end{overprint}
\end{frame}

\begin{frame}[fragile]{A Simple Artist}
\begin{minted}{python}
    class SomeArtist(Artist):
        'An example Artist that implements the draw method'

        def draw(self, renderer):
            # create some objects and use renderer to draw self here
            renderer.draw_path(graphics_context, path, transform)
\end{minted}
\end{frame}

\begin{frame}<presentation:1-|handout:0>{Goals}
    \begin{figure}
        \includegraphics[width=\linewidth]{figures/flow/artists.png}
    \end{figure}
\end{frame}

\begin{frame}<presentation:1|handout:0>{How do we express structure?} 
    \begin{figure}
        \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
        \caption{Figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \end{figure}
\end{frame}

\begin{frame}<presentation:1|handout:0>{Continuity}
    \begin{figure}
        \includegraphics[width=\linewidth]{../paper/figures/k_different_types.png}
    \end{figure}
    \begin{block}{Topological Properties \cite{wilkinsonGrammarGraphics2005}}
        how elements in a dataset are organized, e.g. discrete rows in a table, networked nodes, pixels in an image, points on a line
    \end{block}
\end{frame}

\begin{frame}<presentation:1|handout:0>{Visual Algorithms \& Continuity}
    \begin{figure}
        \includegraphics[width=1\linewidth]{figures/math/plottypes.png}
    \end{figure}
\end{frame}

\begin{frame}<presentation:1-|handout:0>{Equivariance}
    \begin{description}
        \item[What] Retinal Variables \& Marks: visual encodings should match properties of the data \cite{bertinSemiologyGraphicsDiagrams2011a}
        \pause
        \item[Why] Graphical Integrity: graphs show \textbf{only} the data\cite{tufteVisualDisplayQuantitative2001}
        \pause 
        \item[Why] Naturalness: easier to understand when properties match\cite{norman_things_smart}
        \pause
        \item[How] Expressiveness: which structure preserving mappings can a tool implement\cite{mackinlayAutomatingDesignGraphical1986}] 
    \end{description}
\end{frame}

%flip columns
\begin{frame}<presentation:0-|handout:0>{Theoretical Frameworks}
    \begin{description}
        \item[linguistically] visualization has syntax, semantics, grammar expresses how to design structure preserving visualizations
        \cite{mackinlayAutomatingDesignGraphical1986,mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005}

        \item[algebraically] transformations on data and graphics are equivalent and symmetric \cite{kindlmannAlgebraicProcessVisualization2014}
        \begin{columns}
            \column{.5\textwidth}
            \begin{description}
                \item[D] raw data 
                \item[R] representation (data container) 
                \item[V] visualization
            \end{description}
            \column{.5\textwidth}
            \begin{equation*}
                \begin{tikzcd}[ampersand replacement=\&]
                    D \arrow[d, "\alpha"'] \arrow[r, "r_1"] \& R \arrow[r, "\nu"]  \& V \arrow[d, "\omega"] \\
                    D \arrow[r, "r_2"']                     \& R \arrow[r, "\nu"'] \& V                    
                \end{tikzcd}
            \end{equation*}
        \end{columns} 
    %\item[categorically] \textit{understanding} = \textit{read} $\circ$ \textit{render} \cite{vickersUnderstandingVisualizationFormal2013}    
    \end{description}
\end{frame}

\begin{frame}<presentation:1-|handout:1>
    \frametitle{Domain Specific Library: library assumes structure \cite{HeerSoftware2006}}
    \begin{table}
        %\renewcommand{\arraystretch}{2}
        %
        \begin{tabular}{>{\onslide<1->}l>{\onslide<2->}l>{\onslide<3->}l}
            \includegraphics[width=.24\textwidth]{figures/intro/table.png} & \includegraphics[width=.3\textwidth]{figures/intro/landsat.png} & \includegraphics[width=.33\textwidth]{figures/math/graph.png} \\
            ggplot\cite{wickhamGgplot2ElegantGraphics2016a}  & ImageJ\cite{schneiderNIHImageImageJ2012}& Gephi\cite{bastianGephiOpenSource2009}\\
            Vega\cite{satyanarayanDeclarativeInteractionDesign2014} & ImagePlot\cite{studiesCulturevisImageplot2021} & Graphviz\cite{ellsonGraphvizOpenSource2002}\\
            Altair\cite{vanderplasAltairInteractiveStatistical2018}& Napari\cite{nicholas_sofroniew_2021_4533308} & Networkx\cite{HagbergExploringNetwork2008}\\
             Tableau \cite{StoltePolaris2002}& &\\
            \cite{hanrahanVizQL2006,MackinlayShowme2007}&&\\        
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}<presentation:1-|handout:1>
    \frametitle{Building Block Library\cite{wongsuphasawatNavigatingWideWorld2021}: visual algorithms assume structure \cite{toryRethinkingVisualizationHighlevel2004}}
    \begin{figure}
        \includegraphics[height=.3\textheight]{../paper/figures/k_different_types.png}
    \end{figure}

    \begin{enumerate}
        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007} $\rightarrow$ Seaborn\cite{waskom2020seaborn}, xarray \cite{hoyer2017xarray} 
        \item D3 \cite{bostockDataDrivenDocuments2011}
        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012},MayaVi\cite{RamachandranMayaVI2011}$\rightarrow$ Titan\cite{brianwylieUnifiedToolkitInformation2009}, ParaView\cite{ahrens2005paraview}
    \end{enumerate}
\end{frame}

\section{Algebraic Topology & Category Theory}
\begin{frame}<presentation:1|handout:1>{Design Composable Structure Preserving API}
    \begin{description}
        \item[Fiber Bundles] "unified, dimension-independent framework" that expresses data as the mapping between continuity and fields \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}
        \item[Category Theory Language] express constraints in specifications \cite{wielsManagementEvolvingSpecifications1998}   
        \item[Sheaves on Bundles] "algebraic data structure" for representing data over topological spaces \cite{ghristElementaryAppliedTopology2014}
    \end{description}
\end{frame}

\subsection{Fiber Bundles}
\begin{frame}<presentation:1-|handout:1>
    \frametitle<1|handout:1>{Fiber Bundle}
    \frametitle<6|handout:2-3>{Data Bundle}
    \frametitle<7|handout:4-5>{Graphic Bundle}
    \begin{columns}
        \column{0.31\textwidth}
        \begin{equation*}
            %\label{eq:related-work:continuity:fiber-bundle}
            \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
              \onslide*<3-6|handout:1-2>{\dfiberc}
              \onslide*<7|handout:4>{\gfiberc} 
              \onslide*<4-|handout:1-2,4>{\arrow[r, hook]} \& 
              \onslide*<1-6|handout:1-2>{\dtotalc} 
              \onslide*<7|handout:4>{\gtotalc}
              \onslide*<4-|handout:1-2,4>{\arrow[d, "\pi"']} \\
               \& 
               \onslide*<2-6|handout:1-2>{\dbasec}
               \onslide*<7|handout:4>{\gbasec} 
               \onslide*<5-6|handout:1-2>{\arrow[u, "\dsectionc"', bend right, pos=.5]}
               \onslide*<7|handout:4>{\arrow[u, "\gsectionc"', bend right, pos=.5]}
            \end{tikzcd}
        \end{equation*}
     \column{0.69\textwidth}
     \only<1-5|handout:1>{
        \begin{description}
            \item<1-5|handout:1>[\textcolor{total}{Total Space}]{$(\dtotalc, \mathcal{T}_{\dtotalc})$}
            \item<2-5|handout:1>[\textcolor{base}{Base Space}]{$(\dbasec, \mathcal{T}_{\dbasec}),\; \dbasepointc \in \opensetc \subseteq \dbasec$}
            \item<3-5|handout:1>[\textcolor{fiber}{Fiber Space}]{$\dfiberc\restriction_{\dbasepointc} = \pi^{-1}(\dbasepointc), \dfiberc = \dfiberc\restriction_{\dbasepointc}\; \forall \dbasepointc \in \dbasec$}
        \end{description}
        }
    \only<6|handout:2>{
        \begin{description}
            \item[\textcolor{total}{Data \dtotalc}]{continuity + fields}
            \item[\textcolor{base}{Continuity \dbasec}]{how data elements are organized (topological properties) \cite{wilkinsonGrammarGraphics2005}, index (key) space \cite{munznerWhatDataAbstraction2014})}
            \item[\textcolor{fiber}{Fields \dfiberc}] {generalization of a schema - named and typed date fields \cite{spivakSIMPLICIALDATABASES, spivakDatabasesAreCategories2010}}
        \end{description}
        }
        \only<7|handout:4>{
            \begin{description}
                \item[\textcolor{total}{Graphic \gtotalc}]{continuity + renderer fields}
                \item[\textcolor{base}{Continuity \gbasec}]{parameterization of graphic area (e.g. "inked" bounding box\cite{CairographicsOrg})}
                \item[\textcolor{fiber}{Display \gfiberc}] {renderer fields, e.g. \{xy,rgba\}, \{xy, cymk\}, \{xyz, rgba\}}
            \end{description}
            }
    \end{columns}
    \only<4-5|handout:1>{
        \begin{alertblock}{}
            \begin{description}
                \only<5:handout:1>{\item[\textcolor{section}{Sections}]{$\cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\}$}}
            \end{description}
        \end{alertblock}
    }
    \only<4:handout:1>{
        \begin{alertblock}{}
    \begin{description}
    \item[Locally Trivial] {for every point $\dbasepointc \in \dbasec$, there exists an open neighborhood $\dbasepointc \in \opensetc \subseteq \dbasec$ s.t. there is a homeomorphism $\pi^{-1}(\opensetc)\xrightarrow{\equivc} \opensetc \times \dfiberc$}
    \item[(Globally) Trivial]{$\dtotalc = \dbasec \times \dfiberc$}
\end{description}
\end{alertblock}
    }
    \only<6|handout:2>{
    \begin{alertblock}{}
        \begin{description}
            \item[\textcolor{section}{Data} \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}]{$
        \dsectionc(\dbasepointc) = \;\delementc,\; \dbasepointc\in \opensetc \subseteq \dbasec,\; \delementc \in \dfiberc\restriction_{\dbasepointc}$} 
        \item[]{$\delementc=\{field_0:value, \cdots field_i:value, \cdots field_n:value\}$}
        \end{description}
    \end{alertblock}
    }
    \only<7|handout:4>{
        \begin{alertblock}{}
            \begin{description}
                \item[\textcolor{section}{Graphic}]{$\cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}} \coloneqq \big\{\gsectionc: \opensetgc\rightarrow \gtotalc\restriction_{\opensetgc} \; \bigm{\vert} \pi(\gsectionc(\gbasepointc)) = \gbasepointc\;for\, all\; \gbasepointc \in \opensetgc \big\}$}
                \item[] {$\gsectionc(\gbasepointc) = \gelementc,\; \gbasepointc\in \opensetgc\subseteq \gbasec,\; \gelementc \in \dfiberc\restriction_{\gbasepointc}$}
                \item[] $\gelementc=\{x,\,y,\,r,\,g,\,b\}$

        \end{description}    
    \end{alertblock}
    }
    \only<|handout:3>{
        \begin{figure}
            \includegraphics[scale=.65]{figures/math/spherebundle.png}
        \end{figure}
    }
    \only<|handout:5>{
        \begin{figure}
            \includegraphics[scale=.65]{figures/math/render.png}
        \end{figure}
    }
\end{frame}

\subsection{Category Theory Language}
\begin{frame}<presentation:1|handout:1>[fragile]{}
    \begin{minted}{python}
        Artist(Data) -> Graphic
    \end{minted}
\end{frame}

\begin{frame}<presentation:1-|handout:1->
    \frametitle<presentation:1-6|handout:1>{Category \setc}
    \frametitle<presentation:7|handout:2>{Category \setc^{op}}
\begin{columns}
    \column{0.39\textwidth}
    \begin{equation*}
    \begin{tikzcd}[ampersand replacement=\&]
        \onslide*<1-|handout:1>{\setc_1} 
        \onslide*<3-6|handout:1>{\arrow[r, "r_{1,2}", color=set]}
        \onslide*<4-6|handout:1>{\arrow[rd, "r_{1,2}\circ r_{2,3}"', color=set]}
        \onslide*<2-6|handout:1>{\arrow["id_{\setc_1}"', loop, distance=2em, in=215, out=145, color=set]} \& 
        \onslide*<1-|handout:1>{\setc_2}
        \onslide*<7|handout:2>{\arrow[l, color=set]}
        \onslide*<3-6|handout:1>{\arrow[d, "r_{2,3}", color=set]} 
        \onslide*<2-6|handout:1>{\arrow["id_{\setc_2}"', loop, distance=2em, in=125, out=55, color=set]} \\\& 
        \onslide*<1-|handout:1>{\setc_3}
        \onslide*<7|handout:2>{\arrow[u, color=set]}
        \onslide*<7|handout:2>{\arrow[lu, color=set]}
        \onslide*<2-6|handout:1>{\arrow["id_{\setc_3}"', loop, distance=2em, in=305, out=235, color=set]}              
    \end{tikzcd}
    \end{equation*}
    \column{0.59\textwidth}
    \begin{alertblock}<5-6|handout:1>{associativity} 
        if $r_{1,2}: \setc_1 \rightarrow \setc_2$, $r_{2,3}: \setc_2 \rightarrow \setc_3$ and $r_{3,4}: \setc_3 \rightarrow \setc_4$ then $r_{3,4}\circ (r_{2,3} \circ r_{1,2}) = (r_{3,4} \circ r_{2,3}) \circ r_{1,2}$
    \end{alertblock}
    \begin{alertblock}<6|handout:1>{identity} 
        for every $r_{1,2}: \setc_1 \rightarrow \setc_2$ there exists identity morphisms $r_{1,2} \circ id_{\setc_1} = r_{1,2} = id_{\setc_2} \circ r_{1,2}$
    \end{alertblock}
    \end{columns}
\end{frame}

>>>>

%\documentclass[xcolor={dvipsnames}, handout]{beamer}
\documentclass[xcolor={dvipsnames}]{beamer}

%\setbeamertemplate{footline}[frame number]
%\setbeamersize{description width=-\labelsep}
\usetheme{SimpleDarkBlue}

@@

\graphicspath{.figures/, ../papers/figures/}

\usepackage[backend=bibtex, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}

\bibliography{references}

@@

\title{Topologically Equivariant \\ Artist Model}
\author{Hannah Aizenman\texorpdfstring{\\}{,}  Mikael Vejdemo-Johansson \texorpdfstring{\\}{,} Thomas Caswell \texorpdfstring{\\}{,} Michael Grossberg}
\institute{Department of Computer Science, The Graduate Center}
\date{\today}

\begin{document}

\begin{frame}<presentation:1|handout:1>
	\titlepage
    %Hannah Aizenman, Mikael Vejdemo-Johansson, Thomas Caswell, Michael Grossberg\\
    %Committee: Dr. Robert Haralick, Dr. Lev Manovich, Dr. Huy Vo\\
    %External Member: Dr. Marcus Hanwell
\end{frame}

\begin{frame}{About}
    \begin{block}{Me}
        \begin{itemize}
            \item PhD candidate in Computer Science
            \item Matplotlib Community Manager \& Core Developer
            \item  socials/github: story645
        \end{itemize}
    \end{block}

    \begin{block}{Project}
        \begin{itemize}
            \item Funded by Chan Zuckerberg Initiative EOSS 1 \& 3
        \end{itemize}
    \end{block}
\end{frame}

\section{Introduction}
\begin{frame}{Goals}
    \begin{itemize}
        \item uniform API for arbitrary datasets in arbitrary data containers
        \item generalizable methodology for expressing structure preservation
        \item framework that translates into coding specifications
    \end{itemize}
\end{frame}

%% maybe pull out the middle for now? move the 3 stages down to construction?
\begin{frame}{What do visualization libraries do?}
    \begin{figure}
        \begin{overprint}
            \onslide<1|handout:1>\includegraphics[width=\linewidth]{figures/flow/s2.png}
            \onslide<2|handout:2>\includegraphics[width=\linewidth]{figures/flow/s_vc.png}
            \onslide<3|handout:3>\includegraphics[width=\linewidth]{figures/flow/s_mark.png}
            \onslide<4|handout:4>\includegraphics[width=\linewidth]{figures/flow/s3.png}
        \end{overprint}
    \end{figure}
\end{frame}

\begin{frame}{Matplotlib}
    \begin{figure}
       \includegraphics[width=\linewidth]{figures/flow/artists.png}
    \end{figure}
\end{frame}

\begin{frame}{How do we express structure?}

    \begin{description}
        \item[\textcolor{fiber}{\textbf{field}}]is a set of values of the same type, e.g. one column of a table or the pixels of an image
        \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
    \end{description}
\end{frame}

\begin{frame}{Structure Preservation: Equivalent scales and transforms}

\begin{definition}\label{def:related-work:action}\cite{grimaldiDiscreteCombinatorialMathematics2006}
    An \textcolor{action}{\textbf{action}} of \textcolor{action}{$G = (G,\circ, e)$} on $X$ is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.
\end{definition}
\end{frame}

\begin{frame}{Fields: Equivariance }

    Given a group $G$ that acts on both the input $X$ and the output $Y$ of a function $f: X \rightarrow Y$
    \begin{definition}\label{def:equivariance}
     A function $f$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
    \end{definition}
\end{frame}
\begin{frame}{Fields: Equivariance}
    \begin{figure}
        \includegraphics[width=\linewidth]{../paper/figures/equivariant.pdf}
        \label{fig:related-work:equivariance}
      \end{figure}
\end{frame}

\begin{frame}{Fields: Homomorphism}
    Given the function $f: X \rightarrow Y$, with operators $(X, \circ)$ and $(Y, *)$

    \begin{definition}\label{def:homomorphism}
      A function $f$ is \textbf{homomorphic} when $f(x_1 \circ x_2) = f(x_1) * f(x_2)$ and preserves identities $f(I_x) = I_y$ all $x, y \in X$ \cite{grimaldiDiscreteCombinatorialMathematics2006}
    \end{definition}
\end{frame}

\begin{frame}{Fields: Homomorphism}

    \begin{figure}
        \includegraphics[width=\linewidth]{../paper/figures/homomorphism.pdf}
        \label{fig:related-work:homomorphism}
      \end{figure}
\end{frame}

\begin{frame}{Topology: Homeomorphism}

    \begin{definition}
        A function $f$ is a $homeomorphism$ if it is bijective, continuous, and has a continuous inverse function $f^{-1}$.
      \end{definition}
\end{frame}

\begin{frame}{Homeomorphic topology}
    \begin{figure}
        \includegraphics[width=\linewidth]{../paper/figures/k_different_types.pdf}
    \end{figure}
\end{frame}

\begin{frame}{Structure Preservation}
    \begin{itemize}
        \item Bertin: equivalent data and visual field properties \cite{bertinSemiologyGraphicsDiagrams2011}
        \item Mackinlay: homomorphic fields and equivalent topology \cite{mackinlayAutomaticDesignGraphical1987}
        \item Wilkinson: homomorphic scales and  equivalent topology \cite{wilkinsonGrammarGraphics2005}
        \item Kindleman  \& Schieidegger: equivariant scales, invariant data representation, equivalent topology \cite{kindlmannAlgebraicProcessVisualization2014}
    \end{itemize}
    \begin{block}{contribution}
    explicit homeomorphic topology, equivariant actions on fields and topology
    \end{block}
\end{frame}

\begin{frame}{Domain Specific Library: library assumes structure \cite{HeerSoftware2006}}
    \begin{table}
        %\renewcommand{\arraystretch}{2}
        %
        \begin{tabular}{>{\onslide<1->}l>{\onslide<2->}l>{\onslide<3->}l}
            \includegraphics[width=.24\textwidth]{figures/intro/table.png} & \includegraphics[width=.3\textwidth]{figures/intro/landsat.png} & \includegraphics[width=.33\textwidth]{figures/math/graph.png} \\
            ggplot\cite{wickhamGgplot2ElegantGraphics2016}  & ImageJ\cite{schneiderNIHImageImageJ2012}& Gephi\cite{bastianGephiOpenSource2009}\\
            Vega\cite{satyanarayanDeclarativeInteractionDesign2014} & ImagePlot\cite{studiesCulturevisImageplot2021} & Graphviz\cite{ellsonGraphvizOpenSource2002}\\
            Altair\cite{vanderplasAltairInteractiveStatistical2018}& Napari\cite{nicholas_sofroniew_2021_4533308} & Networkx\cite{HagbergExploringNetwork2008}\\
             Tableau \cite{StoltePolaris2002}& &\\
            \cite{hanrahanVizQL2006,MackinlayShowme2007}&&\\
        \end{tabular}
    \end{table}
\end{frame}

\begin{frame}
    \frametitle{Building Block Library\cite{wongsuphasawatNavigatingWideWorld2021}: visual algorithms assume structure \cite{toryRethinkingVisualizationHighlevel2004}}

    \begin{enumerate}
        \item Matplotlib\cite{hunterMatplotlib2DGraphics2007} $\rightarrow$ Seaborn\cite{Waskom2021}, xarray \cite{hoyer2017xarray}
        \item D3 \cite{bostockDataDrivenDocuments2011}
        \item VTK \cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012},MayaVi\cite{RamachandranMayaVI2011}$\rightarrow$ Titan\cite{brianwylieUnifiedToolkitInformation2009}, ParaView\cite{ahrens2005paraview}
    \end{enumerate}
\end{frame}

\section{Algebraic Topology \& Category Theory}
\begin{frame}<presentation:1|handout:1>{Design Composable Structure Preserving API}
    \begin{description}
        \item[Fiber Bundles] Butler: "unified, dimension-independent framework" that expresses data as the mapping between continuity and fields \cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}
        \item[Simplicial Databases] Spivak: Add rich typing for fields to Bulter \cite{spivakSimplicialDatabases2009}
        \item[Category Theory Language] express constraints in specifications \cite{wielsManagementEvolvingSpecifications1998}
        \item[Sheaves] Ghrist "algebraic data structure" for representing data over topological spaces \cite{ghristElementaryAppliedTopology2014}
    \end{description}
\end{frame}

\subsection{Uniform Abstraction for Data \& Graphics}

\begin{frame}<presentation:1|handout:1>{Expressive Types}
    \begin{equation*}
        \textcolor{section}{\texttt{dataset}}: \textcolor{base}{\texttt{topology}} \rightarrow \textcolor{fiber}{\texttt{fields}}
      \end{equation*}
\end{frame}

\subsubsection{Abstract data representation}
\begin{frame}{Fiber Bundle}

    \begin{definition}\label{def:fiber_bundle}
       A \textbf{fiber bundle} $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and  bundle projection map $\pi: \dtotalc \rightarrow \dbasec$ \cite{spanier1989algebraic}.

       \begin{equation} \label{eq:atct:fb:intro}
        \begin{tikzcd}[ampersand replacement=\&]
          \dfiberc \arrow[r, hook, color=total] \& \dtotalc \arrow[r, "\pi", color=total, two heads] \& \dbasec
          \end{tikzcd}
        \end{equation}

    A continuous surjective map $\bm{\pi}$ is a \textbf{bundle projection} map when
    \begin{enumerate}
      \item all fibers in the bundle are isomorphic. Since all fibers are isomorphic $\dfiber \cong \dfiber_{\dbasepoint}$ for all points $\dbasepoint \in \dbase$, there is a uniquely determined \textcolor{fiber}{fiber space} \dfiberc\ given by the preimage of the projection $\pi$ at any point $\dbasepoint$ in the \textcolor{base}{base space} \dbasec: $\dfiberc = \pi^{-1}(k)$.
      \item each point $\dbasepointc$ in the \textcolor{base}{base space} \dbasec\ has an open neighborhood $\openset_{\dbasepointc}$ such that the \textcolor{total}{total space} \dtotalc\ over the neighborhood is locally trivial.
    \end{enumerate}
    \end{definition}

    \textbf{Local triviality} means $\dtotal\vert_{\openset} = \openset\times \dfiber$.
\end{frame}

\begin{frame}{Fiber Bundle}
    \begin{figure}
        \includegraphics[width=\linewidth]{../paper/figures/transition_maps.pdf}
    \end{figure}
\end{frame}

\begin{frame}{Fiber Bundle: section}

    \begin{minipage}{.5\columnwidth}
        \begin{definition}\label{def:fiber_bundle:section}
        A \textcolor{section}{\textbf{section}} $\dsectionc: \dbasec \rightarrow \dtotalc$ over a fiber bundle is a smooth right inverse of $\pi(\dsection(\dbasepoint)) = \dbasepoint$ for all $\dbasepoint \in \dbase$
        \end{definition}
        \end{minipage}
        \begin{minipage}{.4\columnwidth}
          \begin{equation} \label{eq:atct:fb:intro-sec}
            \begin{tikzcd}[ampersand replacement=\&, row sep=huge]
             \dfiberc
              \arrow[r, hook, color=total] \&
              \dtotalc
              \arrow[d, "\pi"',color=total, two heads] \\
               \&
            \dbasec
               \arrow[u, "\dsectionc"', bend right, pos=.5, color=section, dashed]
            \end{tikzcd}
          \end{equation}
        \end{minipage}
\end{frame}

\begin{frame}{Fiber Bundle: section}
    \begin{figure}
        \includegraphics[width=\textwidth]{../paper/figures/fb.pdf}
    \end{figure}
\end{frame}

\subsubsection{Topology: base Space \dbasec}

\begin{frame}{Base space: topology}
The base space of a fiber bundle is a quotient topology\cite{munkresElementsAlgebraicTopology1984}. Given a set $X$ and a function $\mathcal{N}:X\to 2^{2^X}$ that assigns to any $x\in X$ a non-empty collection of subsets $\mathcal{N}(x)$, where each element of $\mathcal{N}(x)$ is a \emph{neighborhood of $x$}, then $X$ with  $\mathcal{N}$ is a \textcolor{base}{topological space} and $\mathcal{N}$ is a neighborhood \emph{topology} if for each $x$ in $X$: \cite{brownronaldTopologyGroupoids2006}

\begin{definition}\label{def:topology}
\begin{enumerate}
  \item if $N$ is a neighborhood $N \in \mathcal{N}(x)$ of $x$ then $x \in N$
  \item every superset of a neighborhood of $x$ is a neighborhood of $x$; therefore a union of a neighborhood and adjacent points in $X$ is also a neighborhood of $x$
  \item the intersection of any two neighborhoods of $x$ is a neighborhood of $x$
  \item any neighborhood $N$ of $x$ contains a neighborhood $M \subset N$ of $x$ such that $N$ is a neighborhood of each of the points in $M$
\end{enumerate}
\end{definition}
\end{frame}

\begin{frame}{Why neighborhood topology?}
\begin{figure}
\includegraphics[width=\linewidth]{../paper/figures/k_different_types.pdf}
\end{figure}
\end{frame}

\begin{frame}{Base \texttt{TYPE} \dbasec}
    \begin{definition}\label{def:atct:category}
        An \textbf{category} $\mathcal{C}$ consists of the following \textit{data}:
     \begin{enumerate}
       \item a collection of \textit{objects} $X \in \textbf{ob}(\mathcal{C})$
       \item for every pair of objects $X, Y \in \textbf{ob}(\mathcal{C})$, a set of \textit{morphisms} $X \xrightarrow{f} Y \in Hom_{\mathcal{C}}(X, Y)$
       \item for every object $X$, a distinct \textit{identity morphism} $X \xrightarrow {id_x} X$ in $Hom_{\mathcal{C}}(X, X)$
       \item a \textit{composition function} $f \in Hom_{\mathcal{C}}(X, Y) \times  g \in Hom_{\mathcal{C}}(Y, Z) \rightarrow g \circ f \in Hom_{\mathcal{C}}(X, Z)$
     \end{enumerate}
     such that
     \begin{enumerate}
       \item \textit{unitality:} for every morphism $ X \xrightarrow{f} Y$, $f \circ id_x = f = id_y \circ f$
       \item \textit{associativity:} if any three morphisms $f, g, h$ are composable,
         \begin{equation*}
           \begin{tikzcd}[ampersand replacement=\&]
             X \arrow[r, "f"] \arrow[rrr, "h\circ(g\circ f) = (h\circ g)\circ f"', bend right, dashed] \& Y  \arrow[r, "g"] \& Z \arrow[r, "h"] \& W
             \end{tikzcd}
       \end{equation*}
       then they are associative such that $h\circ(g\circ f) = (h \circ g) \circ f$  \cite{lawvere2009conceptual,riehlCategoryTheoryContext,maclaneCategoriesWorkingMathematician2013,fongInvitationAppliedCategory2019}.
       \end{enumerate}
     \end{definition}
\end{frame}

\begin{frame}{Add records $\rightarrow$ Add to base space}
    \begin{figure}
        \includegraphics[width=\textwidth]{../paper/figures/tex/k_coproduct.pdf}
    \end{figure}
\end{frame}

\subsubsection{fields: fiber space \dfiberc}

\begin{frame}{Field types }
   A \textit{schema} consists of a pair $(\fnames, \sigma)$ where $\fnames$ is the set of field names and $\sigma: \fnames \rightarrow \textbf{DT}$ is a function from field name to field data type\cite{spivakSimplicialDatabases2009}

   The function $\sigma$ is composed with $\pi$ such that $\pi^{-1}(\sigma(C)) \subseteq \mathscr{U}$; this composition induces a domain bundle $\pi_{\sigma}:\mathscr{U}_{\sigma} \rightarrow \fnames$ that associates a field name $c \in C$ with its corresponding domain $\pi^{-1}_{\sigma}(C) \subseteq \mathscr{U}_{\sigma}$.
